---
title: "Patient Adherence Analysis"
author: "Natalie Youngquist"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# =========================================================================
# 1. LIBRARIES & GLOBAL SETTINGS
# =========================================================================
library(readr)
library(dplyr)
library(caret)
library(pROC)
library(randomForest)
library(gbm)

# =========================================================================
# 2. LOAD DATASET AND TRAINED MODELS
# =========================================================================
patient_train <- readRDS("data/train_data.rds")
patient_test  <- readRDS("data/test_data.rds")

# Load model objects saved from training script
model_lr   <- readRDS("data/lr_model.rds")
model_tree <- readRDS("data/tree_model.rds")
model_rf   <- readRDS("data/rf_model.rds")
model_gbm  <- readRDS("data/gbm_model.rds")

# =========================================================================
# 3. CUSTOM EVALUATION FUNCTIONS
# =========================================================================

# Function to evaluate logistic regression model
evaluate_lr <- function(model, data, outcome_col) {
  actual <- as.integer(data[[outcome_col]])
  probs <- predict(model, newdata = data, type = "response")
  roc_obj <- roc(actual, probs)
  auc_val <- auc(roc_obj)
  pred_class <- ifelse(probs > 0.5, 1, 0)
  
  cm <- confusionMatrix(
    factor(pred_class, levels = c(0,1)),
    factor(actual, levels = c(0,1)),
    positive = "1"
  )
  
  list(
    roc = roc_obj,
    auc = auc_val,
    confusion_matrix = cm,
    predict_class = pred_class
  )
}

# Function to evaluate gradient boosting model
evaluate_gbm <- function(model, data, outcome_col, n.trees) {
  actual <- as.integer(data[[outcome_col]])
  probs <- predict(model, newdata = data, n.trees = n.trees, type = "response")
  roc_obj <- roc(actual, probs)
  auc_val <- auc(roc_obj)
  pred_class <- ifelse(probs > 0.5, 1, 0)
  
  cm <- confusionMatrix(
    factor(pred_class, levels = c(0,1)),
    factor(actual, levels = c(0,1)),
    positive = "1"
  )
  
  list(
    roc = roc_obj,
    auc = auc_val,
    confusion_matrix = cm,
    predict_class = pred_class
  )
}

# Function to plot ROC curve for any model
plot_roc_curve <- function(roc_obj, auc_value, model_name = "Model") {
  plot(
    roc_obj,
    col = "steelblue",
    lwd = 3,
    main = paste("ROC Curve -", model_name)
  )
  
  text(
    x = 0.6,
    y = 0.2,
    labels = paste("AUC =", round(auc_value, 3)),
    col = "steelblue"
  )
}

# Function to calculate the total percentage of correct values across both values
total_percent_correct <- function(actual, predicted_class) {
   percent = sum(actual == predicted_class) / length(actual) * 100
}

# Function to calculate the percentage of predictions by class
percent_correct <- function(actual, predicted) {
  correct_1s <- sum(predicted == 1 & actual == 1)
  total_1s   <- sum(actual == 1)
  pct_1s     <- round((correct_1s / total_1s) * 100, 2)
  
  correct_0s <- sum(predicted == 0 & actual == 0)
  total_0s   <- sum(actual == 0)
  pct_0s     <- round((correct_0s / total_0s) * 100, 2)
  
  data.frame(
    Class = c("1 (Adherent)", "0 (Non-Adherent)"),
    Correct = c(correct_1s, correct_0s),
    Total = c(total_1s, total_0s),
    Percent_Correct = c(pct_1s, pct_0s)
  )
}

# Function to calculate percentage of incorrect predictions by class
percent_incorrect <- function(actual, predicted) {
  incorrect_1s <- sum(predicted == 0 & actual == 1)
  total_1s     <- sum(actual == 1)
  pct_1s       <- round((incorrect_1s / total_1s) * 100, 2)
  
  incorrect_0s <- sum(predicted == 1 & actual == 0)
  total_0s     <- sum(actual == 0)
  pct_0s       <- round((incorrect_0s / total_0s) * 100, 2)
  
  data.frame(
    Class = c("1 (Adherent)", "0 (Non-Adherent)"),
    Incorrect = c(incorrect_1s, incorrect_0s),
    Total = c(total_1s, total_0s),
    Percent_Incorrect = c(pct_1s, pct_0s)
  )
}
```

Before doing any analysis on the models, I wanted to confirm that the data had loaded as expected.

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# confirm the data has loaded as expected
cat("Training Rows:", nrow(patient_train), "\n")
cat("Testing Rows:", nrow(patient_test), "\n")
```

</details>

## Model Evaluation on Training Data

I then fit each model with the training data to measure the response and compared the AUC scores of the Logistic Regression, Random Forest, and Gradient Boosting models.

<details>

<summary><b>Click to expand code</b></summary>

```{r, message=FALSE}
# Logistic AIC Train AUC
pred_train_lr <- predict(model_lr, type = "response")
auc_train_lr <- auc(patient_train$adherent_binary, pred_train_lr)

# Random Forest Train AUC
pred_train_rf <- predict(model_rf, type = "prob")[, 2]
auc_train_rf <- auc(patient_train$adherent_binary, pred_train_rf)

# Gradient Boosting Train AUC
pred_train_gbm <- predict(model_gbm, newdata = patient_train, n.trees = 500, type = "response")
auc_train_gbm <- auc(patient_train$adherent_binary, pred_train_gbm)

# Build Comparison Table
auc_table <- tibble(
  Model = c("Logistic Regression (AIC)", "Random Forest", "Gradient Boosting"),
  AUC   = c(auc_train_lr, auc_train_rf, auc_train_gbm)
)

# compare AUC values
auc_table <- tibble(
  Model = c("Logistic Regression", "Random Forest", "Gradient Boosting"),
  AUC   = c(round(auc_train_lr, digits = 4), round(auc_train_rf, digits = 4), round(auc_train_gbm, digits = 4)
))

auc_table
```

</details>

### Choosing Models

Based on the AUC scores, I chose to continue evaluating the Logistic Regression model, and the Gradient Boosting model. Random Forest performed adequately, but not as well as Gradient Boosting. I continued with Logistic Regression because it is the simpler model and I wanted to see if the complexity from the Gradient Boosting was necessary when compared to the results of the Logistic Regression model.

<details>

<summary><b>Click to expand code</b></summary>

```{r, message=FALSE}
# extract model results on training data
results_lr  <- evaluate_lr(model_lr, patient_train, "adherent_binary")
results_gbm <- evaluate_gbm(model_gbm, patient_train, "adherent_binary", 500)
```

</details>

#### Evaluating Linear Regression Model

I calculated the percentage of correct predictions on the training data using the Logistic Regression model.

The confusion matrix for the Logistic Regression model indicated that:

-   **Correct adherence predictions:** 86.99%

-   **Correct non-adherence predictions:** 61.00%


<details>

<summary><b>Click to expand code</b></summary>

```{r}
# logistic regression performance on training data
#print(results_lr$confusion_matrix)
percent_correct(patient_train$adherent_binary, results_lr$predict_class)
```

</details>

#### Evaluating Gradient Boosting Model

I then calculated the percentage of correct predictions on the training data using the Gradient Boosting model.

The confusion matrix for the Gradient Boosting model indicated that:

-   **Correct adherence predictions:** 90.23%
-   **Correct non-adherence predictions:** 65.53%

The Gradient Boosting model was **3.24%** more accurate at predicting adherent patients and **4.53%%** more accurate at predicting non-adherent patients.


<details>

<summary><b>Click to expand code</b></summary>

```{r}
# gradient boosting performance on training data
#print(results_gbm$confusion_matrix)
percent_correct(patient_train$adherent_binary, results_gbm$predict_class)
```

</details>


#### Visualizing ROC Curve for Both Models

I also plotted the ROC curve to visualize model accuracy on the training data.

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# in sample ROC curves
plot_roc_curve(results_lr$roc, results_lr$auc, "Logistic Regression (Train)")
plot_roc_curve(results_gbm$roc, results_gbm$auc, "Gradient Boosting (Train)")
```

</details>

```{r}
# in sample ROC curves
plot_roc_curve(results_lr$roc, results_lr$auc, "Logistic Regression (Train)")
plot_roc_curve(results_gbm$roc, results_gbm$auc, "Gradient Boosting (Train)")
```

Having performed robust evaluation on the models using the training data, I then moved on to evaluation the models using the testing data.

## Model Evaluation on Testing Data

<details>

<summary><b>Click to expand code</b></summary>

```{r, message=FALSE}
# extract model results on test data
results_lr_test  <- evaluate_lr(model_lr, patient_test, "adherent_binary")
results_gbm_test <- evaluate_gbm(model_gbm, patient_test, "adherent_binary", 500)
```

</details>

#### Evaluating Linear Regression Model

I calculated the percentage of correct predictions on the testing data using the Logistic Regression model.

The confusion matrix for the Logistic Regression model indicated that:

-   **Correct adherence predictions:** 84.66%

-   **Correct non-adherence predictions:** 59.40%

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# logistic regression accuracy on test data
#print(results_lr_test$confusion_matrix)
percent_correct(patient_test$adherent_binary, results_lr_test$predict_class)
```

</details>

#### Evaluating Gradient Boosting Model

I then calculated the percentage of correct predictions on the testing data using the Gradient Boosting model.

The confusion matrix for the Gradient Boosting model indicated that:

-   **Correct adherence predictions:** 88.83%
-   **Correct non-adherence predictions:** 62.08%

The Gradient Boosting model was **4.17%** more accurate at predicting adherent patients and **2.68%%** more accurate at predicting non-adherent patients.

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# gradient boosting accuracy on test data
#print(results_gbm_test$confusion_matrix)
percent_correct(patient_test$adherent_binary, results_gbm_test$predict_class)
```

</details>

#### Percentage of Incorrect Predictions by Each Model

I also wanted to know the percentage of incorrect predictions for each model. I started with Logistic Regression.

The confusion matrix for the Logistic Regression model indicated that:

-   **Incorrect adherence predictions:** 15.34%

-   **Incorrect non-adherence predictions:** 40.60%

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# logistic regression errors on test data
percent_incorrect(patient_test$adherent_binary, results_lr_test$predict_class)
```

</details>


And then evaluated the percentage of incorrect predictions on Gradient Boosting.

The confusion matrix for the Gradient Boosting model indicated that:

-   **Incorrect adherence predictions:** 11.17%

-   **Incorrect non-adherence predictions:** 37.92%

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# overall accuracy on test data
percent_corr_lr  <- total_percent_correct(patient_test$adherent_binary, results_lr_test$predict_class)
percent_corr_gbm <- total_percent_correct(patient_test$adherent_binary, results_gbm_test$predict_class)

cat("Logistic Regression Overall Accuracy:", round(percent_corr_lr, 2), "%\n")
cat("Gradient Boosting Overall Accuracy:", round(percent_corr_gbm, 2), "%\n")
```

</details>


#### Visualizing ROC Curve for Both Models

Lastly, I plotted the ROC curve to visualize model accuracy on the training data.

<details>

<summary><b>Click to expand code</b></summary>

```{r}
# out of sample ROC curves
plot_roc_curve(results_lr_test$roc, results_lr_test$auc, "Logistic Regression (Test)")
plot_roc_curve(results_gbm_test$roc, results_gbm_test$auc, "Gradient Boosting (Test)")
```

</details>

```{r}
# out of sample ROC curves
plot_roc_curve(results_lr_test$roc, results_lr_test$auc, "Logistic Regression (Test)")
plot_roc_curve(results_gbm_test$roc, results_gbm_test$auc, "Gradient Boosting (Test)")
```

### False Non-Adherent Predictions Handling

Predicting a patient is non-adherent when they are actually adherent can lead to significant patient frustration. If an adherent patient experience unresolved ailments or side effects, a correct adherence prediction allows the doctor to rule out inconsistent medication usage as the cause. An incorrect prediction of non-adherence might cause a provider to dismiss these symptoms as a result of poor compliance, damaging the provider-patient trust relationship.

### False Adherent Predictions Handling

Predicting a patient is adherent when they are actually non-adherent is clinically dangerous. It represents a missed opportunity for proactive identification. When a non-adherent patient is flagged as adherent, they do not receive the necessary resources, encouragement, or counseling required to improve their health outcomes, leaving underlying medical issues unaddressed.

## Summary

The choice of model should be dictated by the clinical goal. If the objective is maximum accuracy for a high-risk intervention program, the Gradient Boosting model is superior. If the goal is a cost-effective, explainable baseline that allows providers to clearly see the "why" behind a risk score, Logistic Regression remains a highly effective model.
